Continuous gravitational waves in the lab: recovering audio signals with a table-top optical microphone

### 2020-11-19 paper/ifo-tableTopGWs.tex
minor changes, commented out footmisc
full read-through before submission

% rephrased to mention the laser as the first component and shifted the arm lengths of LIGO and Virgo to later in the para when we discuss them. 
Gravitational-wave detectors such as LIGO and Virgo are large, complex experiments. 
However their design is fundamentally based on the Michelson interferometer, an optical configuration commonly used in undergraduate laboratories. 
In a Michelson interferometer, laser light is split by a beamsplitter into two perpendicular arms (see Fig.~\ref{fig:ifo_schematic_webcam}). 
Mirrors at the end of each arm reflect the two beams back to the beamsplitter where they recombine to produce an interference pattern, 
The resulting interference pattern is dependent on the relative distance travelled by the beams. 
Current generation gravitational-wave observatories have ${\rm km}$-scale arms (with arm lengths $4\,{\rm km}$ and $3\,{\rm km}$ at LIGO and Virgo respectively). 
They are able to measure minuscule changes in distance due to gravitational-waves, e.g. the first detection of a binary black hole merger produced a strain of $10^{-21}$, corresponding to a change in distance which is a fraction of the width of a proton. 
%changed the relative length of the two arms by $10^{-21}$, a distance smaller than the width of a proton~\cite{GW150914}.

\begin{comment}
Gravitational-wave detectors such as LIGO and Virgo are large, complex experiments. However, their design is based fundamentally on a Michelson interferometer, an optical configuration commonly used in undergraduate laboratories.
It consists of a beam splitter which splits light from a laser down two perpendicular arms. % I suggest moving this to later as we are talking about Michelson interferometers generally here. 
%($4\,{\rm km}$ and $3\,{\rm km}$ long at LIGO and Virgo respectively).
Mirrors at the end of each arm reflect the light back towards the beam splitter, where they recombine and produce an interference pattern.
This pattern is dependent on the relative distance travelled by each beam and thus on the relative length of each arm. 
Current generation gravitational-wave observatories are able to measure minuscule changes in distance due to passing gravitational waves. 
The first detection of a binary black hole merger changed the relative length of the two arms by $10^{-21}$~\cite{GW150914}, smaller than the width of a proton.
\end{comment}

% I rephrased to lead with the analogy here and removed 'often' as GW signals are only really listened to for public engagement interest - hannah
Sound is a commonly-used analogy when explaining gravitational-wave science. 
Gravitational-wave signals from binary black hole and neutron star mergers can be converted to audio signals~\footnote{When introducing the acoustic analogy to lay audiences, it is important to emphasize that gravitational waves are not sound. For example, gravitational waves can propagate in a vacuum, whereas sound cannot. Gravitational waves also travel at the speed of light and are not longitudinal waves.} to aid in explanations~\cite{SoundsOfSpaceTime:online,BlackHoleHunter:online}.
% rephrased as this what we are doing, rather than what scientists in general are doing
%This analogue allows scientists to use sound vibrations as a simple means to move the components of a demonstration interferometer. 
Detection and analysis techniques used by the gravitational-wave community can be demonstrated using table-top equipment (see Refs.~\cite{TTExhibit:2020,TTExhibit:online,AMIGO:online,ThorLabsIFO,LIGOIFOGlue,LIGOIFOMagnets,FoxEtAl:1999} for a selection of other table-top Michelson interferometer designs used for science communication).
Sound vibrations provide a simple means to move the components of a demonstration interferometer. 
The vibrations move the mirrors, changing the length of the interferometer arms, and therefore the interference pattern.
We use audio signals throughout this work to simulate gravitational-wave-like signals.
% and  to demonstrate different data analysis and signal processing techniques used to recover such signals.

\begin{figure}
	\includegraphics[width=0.49\textwidth]{figures/ifo_schematic_webcam_edit.pdf}
	\caption{\label{fig:ifo_schematic_webcam}
Schematic of the table-top Michelson interferometer. 
The output of a Class 2 $532\,{\rm nm}$ laser~\cite{ThorLabsIFO} (top-left) is reflected from a mirror at a right angle. The beam then passes through a converging lens (FL = $125.0\,{\rm mm}$) before reaching the beamsplitter which is $50$/$50$ in reflection and transmission. 
The split beams are reflected from mirrors (M1 and M2) at the ends of the interferometer arms and are recombined at the beamsplitter. 
The output interference pattern is enlarged using a diverging lens (FL = $-25.0\,{\rm mm}$), projected onto a screen, and recorded using a commercial webcam connected to a personal computer (PC). 
A speaker fixed to the back of M2 is used to inject audio signals from the PC into the interferometer.
    }
\end{figure}

The optical configuration of the Michelson interferometer used in this work is shown in Fig.~\ref{fig:ifo_schematic_webcam}.
It is assembled on a $450 \times 450 \,{\rm mm} $ optical breadboard and uses a green eye-safe laser (top-left) with peak emission at a wavelength of $532\,{\rm nm}$.
The output of the laser is first reflected by a mirror which turns the beam $90^{\circ}$ (to save space on the optical breadboard and to have greater control of the alignment of the interferometer).
After passing through a converging (plano-convex) lens with focal length $125.0\,{\rm mm}$, the laser beam is incident on a beamsplitter, which reflects half of the light towards mirror 1 (M1) and transmits the other half to mirror 2 (M2). 
A $0.5\,{\rm W}$ speaker is fixed to the back of M2 using commercial adhesive putty and serves as a controllable source of vibrations. 
This speaker is one of a pair of commercial, USB-powered speakers, fed by a $3.5\,{\rm mm}$ jack and driven by a PC. 
The other speaker in the pair is kept face-down and as far away from the interferometer as possible to prevent it from acting as a second source.
The beams are reflected by mirrors M1 and M2, located at the end of $\approx 7.5\,{\rm cm}$ and $\approx 10.0\,{\rm cm}$ long arms.
The beams recombine at the beamsplitter and produce an interference pattern that is then enlarged using a diverging (bi-concave) lens of focal length $-25.0\,{\rm mm}$, and projected onto a screen.
The interference pattern, as shown by the photograph in Fig.~\ref{fig:interference_pattern}, is a set of concentric light and dark rings (fringes). 
A change in the relative length of the arms causes these rings to move radially in or outwards.
Further design details for Michelson interferometers can be found in Ref.~\cite{TTExhibit:2020}.
% Ring interference patterns like this are often used in gravitational-wave table-top demonstrations to allow changes in the pattern to be easily viewed. 
% Why remove the sentence above? Rings are chosen rather than happening by default. - Hannah
The intensity timeseries is recorded via webcam in Sections~\ref{sec:single_tone} and~\ref{sec:viterbi_wandering}.
In Section~\ref{sec:optical_microphone}, which considers more complex audio, the webcam is replaced with a photodiode, which offers a higher sample rate than the webcam. 
%rephrased as this sentence is confusing I think: An intensity time series of some point in the interference pattern is recorded by a webcam (or a photodiode in Section~\ref{sec:optical_microphone}) for data analysis to extract the signal injected at the speaker.

\begin{figure}
 \begin{center}
  \includegraphics[height=.45\textwidth, angle=-90]{figures/webcam_still0_star.pdf}
 \end{center}
 \caption{\label{fig:interference_pattern}
The interference pattern produced by the table-top Michelson interferometer.
This image was taken with the webcam off the beam axis, which may explain why the (predicted spherical) pattern appears elliptical.
The pink star indicates the point where the intensity time series measurements were taken.
The central bright fringe of the interference pattern is approximately $5\,{\rm mm}$ in diameter. 
}
\end{figure}

The motion of the interference fringes follows the motion of M2, and therefore of the speaker.
The amplitude of these motions is given by a transfer function that accounts for the coupling and resonance of the speaker-mirror system.
If the relative change in the length of the arms is kept small enough, the intensity of the interference pattern at any point on the screen oscillates at the same frequency as the injected audio.
But if the relative change in the length of the arms is large enough for multiple fringes to pass over the measured point in a single speaker deformation, then over-counting artificially raises the measured frequency.
As such, any motion of the fringes must be kept small by playing the sound softly through the speaker.
% too much to have --- and () in the same sentence I think 
Even without over-counting, large fringe motions display a nonlinear relation between the intensity fluctuations and injected audio, leading to troublesome---but physically interesting---phenomena like frequency doubling.
We derive this nonlinear relationship between the length difference of the arms and the intensity of the pattern at some point on the screen in Appendix~\ref{app:intensity_derivation}.

>>>>

% rephrased to mention the laser as the first component and shifted the arm lengths of LIGO and Virgo to later in the para when we discuss them. 
Gravitational wave detectors such as LIGO and Virgo are large, complex experiments. 
However their design is fundamentally based on the Michelson interferometer, an optical configuration commonly used in undergraduate laboratories. 
In a Michelson interferometer, laser light is split by a beamsplitter into two perpendicular arms (see Fig.~\ref{fig:ifo_schematic_webcam}). 
Mirrors at the end of each arm reflect the two beams back to the beamsplitter where they recombine to produce an interference pattern, 
The resulting interference pattern is dependent on the relative distance travelled by the beams. 
Current generation gravitational wave observatories have kilometre-scale arms (with arm lengths $4\,{\rm km}$ and $3\,{\rm km}$ at LIGO and Virgo respectively). 
They are able to measure minuscule changes in distance due to gravitational waves, e.g. the first detection of a binary black hole merger produced a strain of $10^{-21}$, corresponding to a change in distance which is a fraction of the width of a proton. 
%changed the relative length of the two arms by $10^{-21}$, a distance smaller than the width of a proton~\cite{GW150914}.

\begin{comment}
Gravitational wave detectors such as LIGO and Virgo are large, complex experiments. However, their design is based fundamentally on a Michelson interferometer, an optical configuration commonly used in undergraduate laboratories.
It consists of a beam splitter which splits light from a laser down two perpendicular arms. % I suggest moving this to later as we are talking about Michelson interferometers generally here. 
%($4\,{\rm km}$ and $3\,{\rm km}$ long at LIGO and Virgo respectively).
Mirrors at the end of each arm reflect the light back towards the beam splitter, where they recombine and produce an interference pattern.
This pattern is dependent on the relative distance travelled by each beam and thus on the relative length of each arm. 
Current generation gravitational wave observatories are able to measure minuscule changes in distance due to passing gravitational waves. 
The first detection of a binary black hole merger changed the relative length of the two arms by $10^{-21}$~\cite{GW150914}, smaller than the width of a proton.
\end{comment}

% I rephrased to lead with the analogy here and removed 'often' as GW signals are only really listened to for public engagement interest - hannah
Sound is a commonly-used analogy when explaining gravitational wave science. 
Gravitational wave signals from binary black hole and neutron star mergers can be converted to audio signals~\footnote{When introducing the acoustic analogy to lay audiences, it is important to emphasize that gravitational waves are not sound. For example, gravitational waves can propagate in a vacuum, whereas sound cannot. Gravitational waves also travel at the speed of light and are not longitudinal waves.} to aid in explanations~\cite{SoundsOfSpaceTime:online,BlackHoleHunter:online}.
% rephrased as this what we are doing, rather than what scientists in general are doing
%This analogue allows scientists to use sound vibrations as a simple means to move the components of a demonstration interferometer. 
Detection and analysis techniques used by the gravitational wave community can be demonstrated using table-top equipment (see Refs.~\cite{TTExhibit:2020,TTExhibit:online,AMIGO:online,ThorLabsIFO,LIGOIFOGlue,LIGOIFOMagnets,FoxEtAl:1999} for a selection of other table-top Michelson interferometer designs used for science communication).
Sound vibrations provide a simple means to move the components of a demonstration interferometer. 
The vibrations move the mirrors, changing the length of the interferometer arms, and therefore the interference pattern.
We use audio signals throughout this work to simulate gravitational wave-like signals.
% and  to demonstrate different data analysis and signal processing techniques used to recover such signals.

\begin{figure}
	\includegraphics[width=0.49\textwidth]{figures/ifo_schematic_webcam_edit.pdf}
	\caption{\label{fig:ifo_schematic_webcam}
Schematic of the table-top Michelson interferometer. 
The output of a Class 2 $532\,{\rm nm}$ laser~\cite{ThorLabsIFO} (top-left) is reflected from a mirror at a right angle. The beam then passes through a converging lens (FL = $125.0\,{\rm mm}$) before reaching the beamsplitter which is $50$/$50$ in reflection and transmission. 
The split beams are reflected from mirrors (M1 and M2) at the ends of the interferometer arms and recombine at the beamsplitter. 
The output interference pattern is enlarged using a diverging lens (FL = $-25.0\,{\rm mm}$), projected onto a screen, and recorded using a commercial webcam connected to a computer (PC). 
A speaker fixed to the back of M2 is used to inject audio signals from the PC into the interferometer.
    }
\end{figure}

The optical configuration of the Michelson interferometer used in this work is shown in Fig.~\ref{fig:ifo_schematic_webcam}.
It is assembled on a $450 \times 450 \,{\rm mm} $ optical breadboard and uses a green eye-safe laser (top-left) with peak emission at a wavelength of $532\,{\rm nm}$.
The output of the laser is first reflected by a mirror which turns the beam $90^{\circ}$ (to save space on the optical breadboard and to have greater control of the alignment of the interferometer).
After passing through a converging (plano-convex) lens with focal length $125.0\,{\rm mm}$, the laser beam is incident on a beamsplitter, which reflects half of the light towards mirror 1 (M1) and transmits the other half to mirror 2 (M2). 
A $0.5\,{\rm W}$ speaker is fixed to the back of M2 using commercial adhesive putty and serves as a controllable source of vibrations. 
This speaker is one of a pair of commercial, USB-powered speakers, fed by a $3.5\,{\rm mm}$ jack and driven by a PC. 
The other speaker in the pair is kept face-down and as far away from the interferometer as possible to prevent it from acting as a second source.
The beams are reflected by mirrors M1 and M2, located at the end of $\approx 7.5\,{\rm cm}$ and $\approx 10.0\,{\rm cm}$ long arms.
The beams recombine at the beamsplitter and produce an interference pattern that is then enlarged using a diverging (bi-concave) lens of focal length $-25.0\,{\rm mm}$, and projected onto a screen.
The interference pattern, as shown by the photograph in Fig.~\ref{fig:interference_pattern}, is a set of concentric light and dark rings (fringes). 
A change in the relative length of the arms causes these rings to move radially in or outwards.
Further design details for Michelson interferometers can be found in Ref.~\cite{TTExhibit:2020}.
% Ring interference patterns like this are often used in gravitational wave table-top demonstrations to allow changes in the pattern to be easily viewed. 
% Why remove the sentence above? Rings are chosen rather than happening by default. - Hannah
The intensity timeseries is recorded via webcam in Sections~\ref{sec:single_tone} and~\ref{sec:viterbi_wandering}.
In Section~\ref{sec:optical_microphone}, which considers more complex audio, the webcam is replaced with a photodiode, which offers a higher sample rate than the webcam. 
%rephrased as this sentence is confusing I think: An intensity time series of some point in the interference pattern is recorded by a webcam (or a photodiode in Section~\ref{sec:optical_microphone}) for data analysis to extract the signal injected at the speaker.

\begin{figure}
 \begin{center}
  \includegraphics[height=0.3\textwidth, angle=-90]{figures/webcam_still0_star.pdf}
 \end{center}
 \caption{\label{fig:interference_pattern}
The interference pattern produced by the table-top Michelson interferometer.
This image was taken with the webcam off the beam axis, which may explain why the (predicted circular) pattern appears elliptical.
The pink star indicates the point where the intensity time series measurements were taken.
The central bright fringe of the interference pattern is approximately $5\,{\rm mm}$ in diameter. 
}
\end{figure}

The motion of the interference fringes follows the motion of M2, and therefore of the speaker.
The amplitude of these motions is given by a transfer function that accounts for the coupling and resonance of the speaker-mirror system.
If the relative change in the length of the arms is kept small enough, then the intensity of the interference pattern at any point on the screen oscillates at the same frequency as the injected audio.
But if the relative change in the length of the arms is large enough for multiple fringes to pass over the measured point in a single speaker deformation, then over-counting artificially raises the measured frequency.
As such, any motion of the fringes must be kept small by playing the sound softly through the speaker.
% too much to have --- and () in the same sentence I think 
Even without over-counting, large fringe motions display a nonlinear relation between the intensity fluctuations and injected audio, leading to troublesome---but physically interesting---phenomena like frequency doubling.
We derive this nonlinear relationship between the length difference of the arms and the intensity of the pattern at some point on the screen in Appendix~\ref{app:intensity_derivation}.

### 2020-11-19 paper/ifo-wanderingTone.tex
full read-through before submission

% the data analysis techniques we draw from 
The analysis techniques used in this section are inspired by those used to search for continuous waves from spinning neutron stars in LIGO and Virgo data~\cite{SuvorovaEtAl:2016,SuvorovaEtAl:2017}. 
Continuous wave searches are performed on long datasets, months to years in duration. 
The frequency of the signal can wander significantly over the observation period. 
In this context \emph{significantly} means across multiple frequency bins, where the typical width of a frequency bin is the reciprocal of the total observation time~\cite{JKS:1998,ScoX1O2Viterbi:2019}.
One method to search for a wandering signal is to split the time series data into several shorter segments which are analysed individually.%, and then recombine them under some continuity assumption to coherently analyse the whole set. 
% I think the following is more accurate. The antenna beam pattern is used to compute a detection statistic. Which detection statistic to use depends on the type of target. -Hannah

A \emph{detection statistic} is used to quantify the likelihood of a signal being present in the data at each frequency and timeseries segment. 
In gravitational-wave data analysis, the detection statistic gives the likelihood of a signal given the antenna beam pattern of the detector, which varies as the Earth rotates and orbits the Sun~\cite{JKS:1998}. 
%Most continuous wave searches use detection statistics to quantify the likelihood of a signal being present in the data at each frequency for each timeseries segment. 
%The detection statistic used depends on the antenna beam pattern of the detector, which varies as the Earth rotates and orbits the Sun, and the type of target~\cite{JKS:1998,SuvorovaEtAl:2017}.
The choice of detection statistic used depends on the type of target~\cite{JKS:1998,SuvorovaEtAl:2017}.
In this work, we take the discrete Fourier transform of each timeseries segment and use its amplitude at each frequency as a detection statistic.
This maximises the likelihood of detecting a sinusoidal signal in Gaussian noise, as described in Appendix~\ref{app:sinusoid_likelihood}. 
The detection statistic is normalised for convenience and we use them to form a grid in  time and frequency (a spectrogram). 
% moved one sentence up 
%We also normalise the detection statistics for convenience.

@@

In Section~\ref{sec:viterbi} we review the Viterbi algorithm and its application to this work. 
In Section~\ref{sec:wanderingResults} we present results for recovery of a wandering audio frequency.

@@

The Viterbi algorithm provides a computationally efficient method for finding the optimal path. 
At every $t_i$ all but $N_f$ possible paths are eliminated (see also Ref.~\cite{ScoX1ViterbiO1:2017}). 
Here we describe the algorithm in pseudocode while referring to the schematic in Fig.~\ref{fig:viterbi}, where the circles indicate the elements in the spectrogram with the lighter colours corresponding to a higher likelihood. The implementation used in this work is available online (see Appendix~\ref{app:code}).
\begin{enumerate}
\item Starting at time $t_1$, each $f_j$ state can originate from three prior states at time $t_0$ (except for $f_0$ and $f_{N_f}$ which only have two). The paths between these states are indicated by the lines in Fig.~\ref{fig:viterbi}. At each $f_j$ state, we select the path with the highest $A[f(t_0),f(t_1)] {\rm Pr}[f(t_0)]$ as the most probable path. These choices are highlighted using the black lines in Fig.~\ref{fig:viterbi} while the grey lines show the rejected paths. For example, the most probable connection to the element labelled (a) is the one directly behind it (i.e., $f_4$). Therefore, this path is selected as the best path from $t_0$ to $t_1$. For backtracking at the end, the index of the most probable connection along with the current cost of the best path is stored for each node.

@@

\begin{figure}
\includegraphics[width=0.49\textwidth]{figures/viterbiDiagram.pdf}
\caption{\label{fig:viterbi}
A schematic diagram of the Viterbi algorithm. 
Each box represents an element in the time-frequency grid starting from the instant $t_0$ on the left and ending at $t_{N_t}$ on the right. 
Vertically, the grid starts from frequency $f_0$ at the bottom and ending at $f_{N_f}$ at the top. 
The colour of each circle indicates the likelihood that a signal is present in the box, where a lighter colour corresponds to higher likelihood. 
At each time, the paths can move up one $f$ bin, move down one $f$ bin, or stay at the same $f$ bin (shown by lines in the diagram). 
Each circle has three paths leading to it indicated by the lines. 
The best path to each circle at each $t_i$ is highlighted in black. 
Some routes through the grid are found to be dead ends, such as the path ending at $t_1$ marked (a). 
At $t=t_{N_t}$ the algorithm chooses the terminating frequency circle which has the highest value given by Eqn.~\ref{eqn:recursiveViterbi}, marked (b). 
The Viterbi path is the path leading to this circle, highlighted in orange from (b) to the start at (c). 
}
\end{figure}

@@

\begin{figure*}
	\includegraphics[width=\textwidth]{figures/expt_overlay_2_viterbi_test_webcam.pdf}
	\caption{\label{fig:viterbi_overlay}
Recovery of a wandering tone. 
The spectrogram shows the observed frequency amplitude in each time-frequency bin. 
The overlaid pink-dot and white-cross markers show the injected signal and recovered Viterbi path respectively. 
On the left, before $\sim 15\,{\rm s}$, the signal changes frequency too quickly for the Viterbi algorithm to recover. 
At $150\,{\rm s}$ the data appears anomalous, which may be due to some background noise. }
\end{figure*}

@@

To simulate a slowly wandering signal we modulate the frequency sinusoidally with a modulation amplitude that decays with time. 
%A sinusoidal signal with slowly decaying amplitude is played through the speaker to mimic a slowly wandering signal. 
We use an identical set up to that shown in Fig.~\ref{fig:ifo_schematic_webcam} and the output interference pattern is recorded via webcam as in Section~\ref{sec:single_tone}. 
%A sinusoidal signal with a slowly decaying amplitude is injected into the interferometer using the speaker in the setup shown in Fig~\ref{fig:ifo_schematic_webcam} to mimic a slowly wandering signal. The particular choice of signal is arbitrary as any injected path would have worked, as long as it changed slowly enough for the search parameters of the algorithm. The output of the interferometer is recorded via a webcam as in Section~\ref{sec:single_tone}. 
In this section we test the Viterbi algorithm’s ability to recover the wandering signal.

The results are shown in Fig.~\ref{fig:viterbi_overlay}, where the heatmap shows the spectrogram of the observed signal. 
For the purposes of this demonstration we use a signal that can easily be identified  in the spectrogram, however real continuous wave signals have much lower amplitudes. 
The overlaid pink dots in Fig.~\ref{fig:viterbi_overlay} show the injected signal and the white crosses show the recovered Viterbi path.% I think this is already defined and seems like repetition:(the most likely path through the data according to the Viterbi algorithm). 
The recovered Viterbi path is within one frequency bin ($\approx 0.3\,{\rm Hz}$) of the injected signal for $94\%$ of the time. We also compute the root mean square (RMS) fractional error $E_{\rm rms}$ along the path which is defined as 
\begin{equation}
E_{\rm rms} = \left[ \frac{1}{N_t+1} \sum_{i=0}^{N_t} \frac{(I_i - R_i)^2}{{I_i}^2} \right] ^{1/2}
\end{equation}
where $I_i$ and $R_i$ are the injected and recovered (Viterbi) frequency paths. Computing this for the result shown in Fig.~\ref{fig:viterbi_overlay}, we find that $E_{\rm rms} = 0.082$, which indicates fair but not total recovery of the injected signal.

>>>>

% the data analysis techniques we draw from 
The analysis techniques used in this section are inspired by those used to search for continuous waves from spinning neutron stars in LIGO and Virgo data~\cite{SuvorovaEtAl:2016,SuvorovaEtAl:2017}. 
Continuous wave searches are performed on long datasets, months to years in duration. 
The frequency of the signal can wander significantly over the observation period. 
In this context, ``significantly'' means across multiple frequency bins, where the typical width of a frequency bin is the reciprocal of the total observation time~\cite{JKS:1998,ScoX1O2Viterbi:2019}.
One method to search for a wandering signal is to split the time series data into several shorter segments which are analysed individually.%, and then recombine them under some continuity assumption to coherently analyse the whole set. 
% I think the following is more accurate. The antenna beam pattern is used to compute a detection statistic. Which detection statistic to use depends on the type of target. -Hannah

A ``detection statistic'' is used to quantify the likelihood of a signal being present in the data at each frequency and timeseries segment. 
In gravitational wave data analysis, the detection statistic gives the likelihood of a signal given the antenna beam pattern of the detector, which varies as the Earth rotates and orbits the Sun~\cite{JKS:1998}. 
%Most continuous wave searches use detection statistics to quantify the likelihood of a signal being present in the data at each frequency for each timeseries segment. 
%The detection statistic used depends on the antenna beam pattern of the detector, which varies as the Earth rotates and orbits the Sun, and the type of target~\cite{JKS:1998,SuvorovaEtAl:2017}.
The choice of detection statistic used depends on the type of target~\cite{JKS:1998,SuvorovaEtAl:2017}.
In this work, we take the discrete Fourier transform of each timeseries segment and use its amplitude at each frequency as a detection statistic.
This maximises the likelihood of detecting a sinusoidal signal in Gaussian noise, as described in Appendix~\ref{app:sinusoid_likelihood}. 
The detection statistic is normalised for convenience and we use it to form a grid in time and frequency (a spectrogram) which displays the frequency content of the signal over time. 
% moved one sentence up 
%We also normalise the detection statistics for convenience.

@@

In Section~\ref{sec:viterbi} we review the Viterbi algorithm and its application to this work. 
In Section~\ref{sec:wanderingResults} we present results for recovery of a wandering frequency audio signal.

@@

The Viterbi algorithm provides a computationally efficient method for finding the optimal path. 
At every $t_i$ all but $N_f$ possible paths are eliminated (see also Ref.~\cite{ScoX1ViterbiO1:2017}). 
Here we describe the algorithm in pseudocode while referring to the schematic in Fig.~\ref{fig:viterbi}, where the circles indicate the elements in the spectrogram with the lighter colours corresponding to a higher likelihood. The implementation used in this work is available online (see Appendix~\ref{app:code}).
\begin{enumerate}
\item Starting at time $t_1$, each $f_j$ state can originate from three prior states at time $t_0$ (except for $f_0$ and $f_{N_f}$ which only have two). The paths between these states are indicated by the lines in Fig.~\ref{fig:viterbi}. At each $f_j$ state, we select the path with the highest $A[f(t_0),f(t_1)] {\rm Pr}[f(t_0)]$ value as the most probable path. These choices are highlighted using the black lines in Fig.~\ref{fig:viterbi} while the grey lines show the rejected paths. For example, the most probable connection to the element labelled (a) is the one directly behind it (i.e., $f_4$). Therefore, this path is selected as the best path from $t_0$ to $t_1$. To allow backtracking at the end, the index of the most probable connection to a node along with the value of the best path to that node is stored, for each node.

@@

\begin{figure}
\includegraphics[width=0.49\textwidth]{figures/viterbiDiagram.pdf}
\caption{\label{fig:viterbi}
A schematic diagram of the Viterbi algorithm. 
Each circle represents an element in the time-frequency grid starting from the instant $t_0$ on the left and ending at $t_{N_t}$ on the right. 
Vertically, the grid starts from frequency $f_0$ at the bottom and ending at $f_{N_f}$ at the top. 
The colour of each circle indicates the likelihood that a signal is present in the box, where a lighter colour corresponds to higher likelihood. 
At each time, the paths can move up one $f$ bin, move down one $f$ bin, or stay at the same $f$ bin (shown by lines in the diagram). 
Each circle has three paths leading to it, indicated by the lines. 
The best path to each circle at each $t_i$ is highlighted in black. 
Some routes through the grid are found to be dead ends, with respect to optimally reaching the other side, such as the path ending at $t_1$ marked (a). 
At $t=t_{N_t}$ the algorithm chooses the terminating frequency circle which has the highest value given by Eqn.~\ref{eqn:recursiveViterbi}, marked (b). 
The Viterbi path is the path leading to this circle, highlighted in orange from (b) to the start at (c). 
}
\end{figure}

@@

\begin{figure*}
	\includegraphics[width=\textwidth]{figures/expt_overlay_2_viterbi_test_webcam.pdf}
	\caption{\label{fig:viterbi_overlay}
Recovery of a wandering tone using the Viterbi algorithm. 
The spectrogram shows the observed frequency (Fourier) amplitude in each time-frequency bin. 
The overlaid pink-dot and white-cross markers show the injected signal and recovered Viterbi path, respectively. 
On the left, before $\sim 15\,{\rm s}$, the signal changes frequency too quickly for the Viterbi algorithm to recover, given a parameter for the rate of change of one frequency bin per time bin. 
At $150\,{\rm s}$ the data appears anomalous, which may be due to some background noise. }
\end{figure*}

@@

To simulate a slowly wandering signal we modulate the frequency sinusoidally with a modulation amplitude that decays with time. 
%A sinusoidal signal with slowly decaying amplitude is played through the speaker to mimic a slowly wandering signal. 
We use the same set-up as shown in Fig.~\ref{fig:ifo_schematic_webcam} and the output interference pattern is recorded via webcam as in Section~\ref{sec:single_tone}. 
%A sinusoidal signal with a slowly decaying amplitude is injected into the interferometer using the speaker in the setup shown in Fig~\ref{fig:ifo_schematic_webcam} to mimic a slowly wandering signal. The particular choice of signal is arbitrary as any injected path would have worked, as long as it changed slowly enough for the search parameters of the algorithm. The output of the interferometer is recorded via a webcam as in Section~\ref{sec:single_tone}. 
In this section we test the Viterbi algorithm’s ability to recover the wandering signal.

The results are shown in Fig.~\ref{fig:viterbi_overlay}, where the heatmap shows the spectrogram of the observed signal. 
For the purposes of this demonstration we use a signal that can easily be identified in the spectrogram, however, we expect real continuous wave signals to have far lower signal-to-noise ratios (SNR's). 
The overlaid pink dots in Fig.~\ref{fig:viterbi_overlay} show the injected signal and the white crosses show the recovered Viterbi path.
% I think this is already defined and seems like repetition:(the most likely path through the data according to the Viterbi algorithm). 
The recovered Viterbi path is within one frequency bin ($\approx 0.3\,{\rm Hz}$) of the injected signal for $94\%$ of the time. We also compute the root-mean-square (RMS) fractional error $E_{\rm rms}$ along the path, which is defined as 
\begin{equation}
E_{\rm rms} = \left[ \frac{1}{N_t+1} \sum_{i=0}^{N_t} \frac{(I_i - R_i)^2}{{I_i}^2} \right] ^{1/2}
\end{equation}
where $I_i$ and $R_i$ are the injected and recovered (Viterbi) frequency paths. Computing this for the result shown in Fig.~\ref{fig:viterbi_overlay}, we find that $E_{\rm rms} = 0.082$, which indicates fair but not total recovery of the injected signal.
